Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-06-26 doc/manuscript_hkk_drought_sensitivity.Rmd
changes to intro

Intense droughts are becoming more frequent, affecting the demography, diversity, and  carbon (C) cycling of forests globally [@ipcc_climate_2023]. 
However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, studies have separately shown that drought characteristics, species traits, and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought. 
In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove vulnerable to a drought with different characteristics. 
The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of climatic anomalies [@slette_how_2019] with potentially distinct ecological consequences.
<!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different seasonal timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry-seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry-season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
Tree species adaptations to drought vary widely, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
In tropical seasonal forests in particular, trees have evolved to optimize growth, reproduction, and survival (i.e., fitness) under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry-season to avoid foliar water loss [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry-season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
<!-- Finally, the microenvironments that individual trees experience influence growth responses to drought leading to varied atmospheric and hydrological stresses with distinct ecological consequences. -->
Microenvironments defined by horizontal topographic and edaphic variation, and vertical light and thermal variation further shape individual tree experience of, and response to, drought. 
While habitats with relatively moist soils (e.g., low-lying topography) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted (e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021).
Indeed, during drought trees near streams experienced greater growth declines in a temperate forest [@mcgregor_tree_2021] and increased mortality in a tropical forest [@zuleta_droughtinduced_2017]. 
<!-- Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments. -->
Tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
These aboveground gradients in crown exposure could be partially offset by the tendency for large trees to have larger/deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and therefore the potential to access more reliable water sources [@stahl_depth_2013]. 
<!-- Sean::: I need to look at this source. Is this true? it doesn't really make sense unless coupled with other traits (e.g., deciduousness shifts accross ontogeny) as 'putting down a tap root' is one of the first things a seedling of a deep rooted species does (i.e., it's not allometry but strategy) -->
However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcys_2015; @fernández-de-uña_role_2023], or by trait differences across vertical axis [@vinod_thermal_2023].
Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.

@@

<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and microenvironments, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats. 
Specifically, we test predictions that:
i) both species and individuals respond differently to different droughts;
ii) deciduous species, which rely on wet-seasons for growth, are particularly sensitive to wet-season drought; and,
iii) exposed trees are more sensitive to drought, but water availability, either through larger topographic location or tree size (and corresponding rooting volume), can buffer growth declines.
We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 
During this period, the forest experienced three different dry periods, two associated with ENSO events, and all three differing with respect to timing and duration (Table 1). 
With data on 1820 trees across 30 species we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
We used a causal modelling framework to account for potential confounding relationships among the variables. 
This analysis adds important insight into the complexity of seasonal dry forest response to drought in species-rich tropical forests.

>>>>

Intense droughts are becoming more frequent [@ipcc_climate_2023], affecting the demography, diversity, and  carbon (C) cycling of forests globally. 
However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict [@evans_scaling_2025], particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, studies have separately shown that drought characteristics, species traits, and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought. 
In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove vulnerable to a drought with different characteristics. 
The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of climatic anomalies [@slette_how_2019].
<!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], differences in seasonal timing, duration, or intensity of droughts may result in variable tree growth responses, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry-season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
Tree species adaptations to drought vary widely, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
In tropical seasonal forests in particular, trees have evolved to optimize growth, reproduction, and survival (i.e., fitness) under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry-season to avoid foliar water loss [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry-season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate $\pi_{tlp}$, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
<!-- Finally, the microenvironments that individual trees experience influence growth responses to drought leading to varied atmospheric and hydrological stresses with distinct ecological consequences. -->
Microenvironments defined by horizontal topographic and edaphic variation and vertical light and thermal variation further shape individual tree experience of, and response to, drought. 
While habitats with relatively moist soils (e.g., low-lying topography) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021].
Indeed, during drought trees near streams experienced greater growth declines in a temperate forest [@mcgregor_tree_2021] and increased mortality in a tropical forest [@zuleta_droughtinduced_2017]. 
<!-- Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments. -->
Tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
These aboveground gradients in crown exposure could be partially offset by the tendency for large trees to have larger/deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and therefore the potential to access more reliable water sources [@stahl_depth_2013]. 
<!-- Sean::: I need to look at this source. Is this true? it doesn't really make sense unless coupled with other traits (e.g., deciduousness shifts accross ontogeny) as 'putting down a tap root' is one of the first things a seedling of a deep rooted species does (i.e., it's not allometry but strategy) -->
However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], this may be driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcys_2015; @fernández-de-uña_role_2023], or by trait differences across vertical axis [@vinod_thermal_2023].
Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.

@@

<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and microenvironments, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats. 
Specifically, we test predictions that:
i) both species and individuals respond differently to different droughts;
ii) deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet-season drought; and,
iii) exposed trees are more sensitive to drought, whereas water access, either through topographic location or tree size (and corresponding rooting volume), has variable effects.
We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 
During this period, the forest experienced three different dry periods, two associated with ENSO events, and all three differing with respect to timing and duration (Table 1). 
With data on 1820 trees across 30 species we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
We used a causal modelling framework to account for potential confounding relationships among the variables. 
This analysis adds important insight into the complexity of seasonal dry forest response to drought in species-rich tropical forests.
